Demes: a standard format for demographic models

### 2022-04-27 paper.tex
some edits and de-duplication

\ggcomment{We don't showcase Demes output from any inference software.}
\end{abstract}

@@

\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-parsers}}

While the high-level specification is designed for human readability and
conciseness, the underlying data model suitable for software implementation
is redundant and exhaustive. 
Translation from the high-level specification to the low-level data model
requires resolving hierarchically-defined default values and verifying
relationships between populations and the validity of specified parameter
values.
Because this translation and validation requires significant
programming effort, we define a standard software entity as part of the
specification to perform this task (the parser),
which is intended to be shared by programs that support Demes as an input.
By having relatively few high-quality Demes parsers available as libraries
(ideally one per programming language), we ensure consistency across simulation
software in their implementation of Demes-specified models.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs.
In addition, we have high-quality parser implementations in the Python and
C languages (published under liberal open source licences), providing a
solid foundation for the software ecosystem (Table~\ref{tab:software}).

@@

\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-scope}}

The Demes specification is static by design---we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.

\aprcomment{I think this belongs here, just one or two sentences:}

\krtcomment{Here, we should say something about other models that people may wish to implement within the demes
framework.}

\krtcomment{
    I think we want a section in the Discussion outlining some obvious things
    that people can do w/demes that are sitting "just outside" the strict
    specification: hard selection, etc..
}

Demes is also designed to interface with a large number of simulation and
inference methods, so we have restricted the specification to only
describe demographic features at the population level.
The underlying discrete demographic structure is the common thread connecting
many widely-used population genetic methods, so we have omitted many
features of the underlying genome biology (as described above, and in
Appendix~\ref{sec:appendix-scope}).

@@

%\krtcomment{The paragraph(s) below subsume(s)/replace(s) some of what appears in later pargraphs.
%	I'm not making an attempt yet to revise that later content--I think other authors should
%	weigh in.}

%\krtcomment{Aaron -- I have perhaps over-simplified the description of marginal allele frequency dynamics?}

@@

% \krtcomment{
%     We need to revisit the paragraph(s) below.
%     We do not assume individuals are exchangeable, because we do not
% 	require that all individuals have the same fitness.
%     See, e.g., p 60 of Wakeley for why fitness == exchangeability
% 	and why that is required for the standard coalescent.
% }
% \mhcomment{The `exchangeability' assumption also breaks down under 
% 	different reproductive modes, since the sampling of the alleles
%     affects coalescent times. 
% 	For example, in a highly selfing species the coalescent time for two different alleles 
% 	within an individual is much lower than two alleles from different individuals.
%     I've triedto rewrite the below paragraph to update and clarify
%     the reproduction parameters in line with the discussions on Github.
%     Finally, 'rates' imply events occur as Poisson-like 
% 	processes over continuous time, whereas the values used here instead define the 
% 	\emph{fraction} of reproductions of each type.
%     Shall we rename to selfing/clonal `fractions',
% 	or would that mess up the rest of the schema too much?
% }

@@

By allowing the definition of selfing and cloning probabilities, we allow many 
standard models to be defined.
However, by parameterising selfing and cloning as we have, we assume that these
properties of populations can be specified independently from the genetics.
In other words, mutations that cause selfing probabilities to fluctuate within
an epoch are not considered.
More details of the mathematical properties of selfing and cloning rates
in a coalescent context can be found in \citet{hartfield2016facsexcoal}.

\mhcomment{If requested
I can quickly draw up a schematic diagram to illustrate these partitions.}

@@

% https://github.com/popsim-consortium/demes-paper/issues/3

The Demes specification is static by design--we wish to
unambiguously describe a demographic model defined by a concrete set
of parameters. This simplicity means that we cannot directly
specify parameter distributions or estimated confidence intervals
for those parameters. While it is not difficult to imagine extending
the specification in ways that would allow this, this would greatly
increase parser complexity and require limiting choices to be made.
Instead, we have opted to restrict Demes to static model descriptions,
although allowing parameterized distributions or parameter ranges 
is an direction of possible future work.

%\krtcomment{
%    Aaron has done a bit of this non-static stuff in moments.
%    It may be useful to look over there?
%}

%\ggcomment{
%Possible reasons to advocate for non-static features are as follows.
%\begin{enumerate}
%\item Specifying prior distributions of parameters for an ABC.
%\item Specifying upper and lower bounds of parameters and inter-parameter
%      constraints for input to an optimisation-based inference.
%\item Specifying joint distribution, or confidence intervals, for the output of
%      an inference procedure.
%\item Simulating from some arbitrary joint distribution that was inferred
%      previously. E.g. using an ABC posterior distribution of models to
%      train a neural network.
%\item Extending someone else's static model to reinfer parameters,
%      or add new parameters.
%\end{enumerate}
%}

@@

\begin{figure}[h!]
    \begin{minipage}{0.44\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
        \end{tcolorbox}
        \begin{tcolorbox}
            \includegraphics[width=\textwidth]{fig/IM}
        \end{tcolorbox}
    \end{minipage}\hfill
    \begin{minipage}{0.54\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
        \end{tcolorbox}
    \end{minipage}\\
    \caption{
        \label{fig:IM}
        Example isolation-with-migration Demes model. (A) The compact non-redundant
        representation expressed using YAML. (B) The same model in the fully-qualified
        form. (C) A visual representation of the model using \texttt{demesdraw}.
        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
        but are much more human-readable.
    }
\end{figure}

\begin{figure}[h!]
    \begin{tcolorbox}
        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yaml}
    \end{tcolorbox}
    \caption{
        \textbf{The \citet{tennessen2012evolution} two-population demographic model in Demes format.}
    }
    \label{fig:tennessen}
\end{figure}

>>>>

\ggcomment{We don't showcase Demes output from any inference software.}
\aprcomment{You could say that ``showcase'' also refers to the table that
includes demes support in inference software, not just the illustration
in the main text.}
\end{abstract}

@@

While the high-level specification is designed for human readability and
conciseness, the underlying data model suitable for software implementation
is redundant and exhaustive. 
Translation from the high-level specification to the low-level data model
requires resolving hierarchically-defined default values and verifying
relationships between populations and the validity of specified parameter
values.
Because this translation and validation requires significant
programming effort, we define a standard software entity as part of the
specification to perform this task (the parser),
which is intended to be shared by programs that support Demes as an input.
By maintaining high-quality Demes parsers available as libraries,
we ensure consistency across simulation
software in their implementation of Demes-specified models.
In addition to a reference implementation written in Python,
we provide high-quality parser implementations in the Python and
C languages (Table~\ref{tab:software}, and discussed in
Appendix~\ref{sec:appendix-parsers}).

@@

The Demes specification is static by design---we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.
It is important to note, however, that Demes may be used in applications that
include additional population genetic processes outside of what is explicitly
modelled in the specification, such as interpreting population sizes as
carrying capacities, implementations of \emph{hard} selection, or layering more
complicated mating or spatial structure. The Demes specification therefore
provides a basic model that can be elaborated on where necessary.

Demes is also designed to interface with a large number of simulation and
inference methods, so we have restricted the specification to only
describe demographic features at the population level.
The underlying discrete demographic structure is the common thread connecting
many widely-used population genetic methods, so we have omitted many
features of the underlying genome biology (Appendix~\ref{sec:appendix-scope}).

@@

@@

@@

By allowing the definition of selfing and cloning probabilities, we allow many 
standard models to be defined.
However, by parameterising selfing and cloning as we have, we assume that these
properties of populations can be specified independently from the genetics.
In other words, mutations that cause selfing probabilities to fluctuate within
an epoch are not considered.
More details of the mathematical properties of selfing and cloning rates
in a coalescent context can be found in
\citet{nordborg1997coalescent,hartfield2016facsexcoal}.

%\mhcomment{If requested
%I can quickly draw up a schematic diagram to illustrate these partitions.}

@@

The Demes specification is designed to describe demographic
models defined by a fixed set of model parameters.
As described in the main text, it does not include
information about estimated confidence intervals or the joint distribution
of parameter values.
While we have opted to restrict Demes to static model descriptions,
allowing parameterized distributions or parameter ranges 
is a direction of possible future work.

@@

\begin{figure}[h!]
    \begin{minipage}{0.44\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
        \end{tcolorbox}
        \begin{tcolorbox}
            \includegraphics[width=\textwidth]{fig/IM}
        \end{tcolorbox}
    \end{minipage}\hfill
    \begin{minipage}{0.54\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
        \end{tcolorbox}
    \end{minipage}\\
    \caption{
        \label{fig:IM}
        Example isolation-with-migration Demes model. (A) The compact non-redundant
        representation expressed using YAML. (B) The same model in the fully-qualified
        form. (C) A visual representation of the model using \texttt{demesdraw}.
        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
        but are much more human-readable.
        \aprcomment{Need figure panel labels.}
    }
\end{figure}

\begin{figure}[h!]
    \begin{tcolorbox}
        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yaml}
    \end{tcolorbox}
    \caption{
        \textbf{The \citet{tennessen2012evolution} two-population
        demographic model in Demes format.}
        This model includes a single ancestral population that expands in
        size in the past, followed by divergence between AFR- and EUR-labeled
        populations. The two-population phase of the model includes multiple
        epochs of varying size, and rapid exponential growth over the past
        five thousand years in each population.
    }
    \label{fig:tennessen}
\end{figure}
